---
title: "Introduction to R Programming"
author: "Dhruv Nawani"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
# Global chunk options
# echo = TRUE      : show R code
# message = FALSE : suppress messages
# warning = FALSE : suppress warnings
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Introduction

R is a programming language and environment widely used for:
- Statistical analysis
- Data visualization
- Data science and machine learning
- Academic and industry research

This document introduces the **fundamental concepts of R** using simple and reproducible examples.

---

# Using Built-in Data

R includes several built-in datasets for learning and experimentation.

The `lynx` dataset records annual numbers of lynx trappings in Canada.

```{r}
hist(lynx,
     main = "Histogram of Lynx Trappings",
     col = "lightblue",
     xlab = "Number of Lynx")
```

**Key idea:**  
A histogram is used to understand the distribution of a numeric variable.

---

# Basic Arithmetic in R

R supports all standard arithmetic operations.

```{r}
12 + 3          # Addition
(12 / 3) * 3    # Division and multiplication
(45 / 5) * 12 + 10
(23 * 3) - (43 * 2)
```

---

## Assignment Operator

Values are stored in variables using the assignment operator `<-`.

```{r}
X <- (45 / 5) * 12 + 10
Y <- (23 * 3) - (43 * 2)

X - Y
```

**Note:**  
Although `=` can be used, `<-` is the conventional assignment operator in R.

---

# Getting Help and Documentation in R

R provides extensive **built-in documentation** for all functions, datasets, and packages.
This documentation can be accessed directly from **RStudio**.

---

## Using the `?` Operator

The question mark `?` is a shortcut to access help files.

```{r}
?hist
```

This opens the help page for the `hist()` function in the **Help** pane of RStudio.

---

## Using the `help()` Function

The `help()` function performs the same task as the `?` operator.

```{r}
help(mean)
```

---

## Getting Help for Datasets

You can also view documentation for built-in datasets.

```{r}
?iris
```

---

## Searching for Help When You Are Unsure

If you are not sure about the exact function name, use `help.search()`.

```{r}
help.search("regression")
```

---

## Why This Is Important

Reading the documentation helps you understand:
- What a function does
- What arguments it takes
- What values it returns
- Examples of correct usage

Learning to use R’s help system is a key step toward becoming independent in R programming.

# Variables and Vectors

A **vector** is the most basic data structure in R.  
All elements in a vector must be of the same data type.

```{r}
PARatings <- c(3, 2, 4, 3, 5, 4, 3, 2, 3, 4, 5)

Name <- c("Ravi", "Sudha", "Ramesh", "Soham", "Shilpa",
          "Bhavesh", "Amarpreet", "Wilfred", "Abdul", "Srini", "Taposh")

Age <- c(23, 22, 23, 25, 26, 24, 27, 23, 23, 25, 26)
```

---

# Data Frames

A **data frame** is a tabular data structure where:
- Each column is a vector
- Each row is an observation

```{r}
PerfData <- data.frame(Name, PARatings, Age)
PerfData
```

---

## Inspecting a Data Frame

```{r}
dim(PerfData)      # Dimensions (rows, columns)
str(PerfData)      # Structure of the data
summary(PerfData)  # Summary statistics
```

These functions are essential for initial data exploration.

---

# Working with Files (Reproducible Approach)

Avoid using `setwd()` as it makes code non-reproducible.  
Instead, read and write files relative to the project directory.

```{r}
write.csv(PerfData, "Appraisal.csv", row.names = FALSE)
PerfData1 <- read.csv("Appraisal.csv")
```

```{r}
str(PerfData1)
summary(PerfData1)
```

---

# Built-in Dataset: Iris

The `iris` dataset is commonly used for demonstrating classification and visualization techniques.

```{r}
data(iris)
str(iris)
summary(iris)
```

```{r}
head(iris)
tail(iris)
```

---

# Visualizing Categorical Data

Categorical (qualitative) data represents groups or categories.

## Bar Plot

```{r}
tab <- table(iris$Species)

barplot(tab,
        main = "Species Distribution",
        xlab = "Species",
        ylab = "Frequency",
        col = rainbow(3))
```

---

## Pie Chart

```{r}
pie(tab,
    main = "Species Composition",
    col = rainbow(3))
```

Bar plots and pie charts are commonly used to show category frequencies.

---

# Visualizing Numerical Data

## Histogram

```{r}
hist(iris$Sepal.Length,
     main = "Histogram of Sepal Length",
     xlab = "Sepal Length",
     col = "orange",
     breaks = 15)
```

Histograms show the shape, center, and spread of numeric data.

---

## Scatter Plot

```{r}
plot(iris$Sepal.Length, iris$Sepal.Width,
     col = "red",
     main = "Sepal Length vs Sepal Width",
     xlab = "Sepal Length",
     ylab = "Sepal Width")
```

Scatter plots visualize relationships between two numeric variables.

---

## Multiple Scatter Plots and Correlations

```{r}
pairs(iris)
```

```{r}
library(psych)
pairs.panels(iris)
```

These plots help identify correlations among multiple variables.

---

# Data Types in R

R commonly works with numeric, character, and logical data types.

```{r}
A <- 4
class(A)

B <- "gender"
class(B)

C <- TRUE
class(C)
```

---

# Vectors, Coercion, and Lists

## Named Vectors

```{r}
Age <- c(Ravi = 23, Sudha = 22, Ramesh = 23, Soham = 25, Shilpa = 26)
Age
```

---

## Type Coercion

When different data types are combined in a vector, R coerces them to a common type.

```{r}
mixed <- c(23, 22, "Ravi", 25)
class(mixed)
```

Character data has higher priority than numeric data.

---

## Lists

Lists can store elements of different data types.

```{r}
mylist <- list(23, 22, "Ravi", TRUE)
sapply(mylist, class)
```

---

# Factors (Categorical Variables)

Factors are used to represent categorical and ordinal data.

```{r}
PARatings <- c(3, 2, 4, 3, 5, 4, 3, 2, 3, 4, 5)

PARatings_factor <- factor(PARatings)

PARatings_ordered <- factor(PARatings,
                            ordered = TRUE,
                            levels = c(2, 3, 4, 5))

str(PARatings_ordered)
```

Ordered factors are useful when category levels have a natural order.

---

# Conditional Statements

Conditional statements allow R to make decisions based on logical conditions.

```{r}
x <- 10

if (x > 0) {
  "x is positive"
} else {
  "x is negative"
}
```

```{r}
if (iris$Sepal.Length[1] < 4) {
  "Sepal length is less than 4"
} else {
  "Sepal length is greater than or equal to 4"
}
```

---

# Operators in R

## Arithmetic Operators

```{r}
A <- 12
B <- 10

A * B
A - B
```

---

## Relational Operators

```{r}
A < B
A > B
A == B
A != B
```

---

## Logical Operators

```{r}
p <- 10

(p < 15 & p > 8)   # AND
(p > 15 | p > 8)   # OR
!(p < 15)          # NOT
```

---

# Conclusion

This document introduced the foundational concepts of R programming, including:
- Data structures
- Visualization techniques
- File handling
- Conditional logic and operators

These fundamentals form the basis for advanced topics such as statistics, data analysis, and machine learning.
